Untitled

### 2018-04-11 April_Presentation.Rmd
adding a ton of stuff to the repo

---
title: "Untitled"
author: "Paul Harmon"
date: "April 2, 2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

The Carnegie Classifications delineate doctoral-granting institutions into three groups: 
+ R1: Very High Research
+ R2: High Research
+ R3: Moderate Research

@@

@@

+ Rank Institutions
+ Index Creation: PCA of aggregate and per-capita variables
+ Plot the Indices (per-capita vs aggregate)
+ Create groups (via line-drawing)

## Principal Components Analysis 
Principal Componenents Analysis is a method for dimension reduction that takes a set of p predictor variables and decomposes it into k principal components that explain the most variation in the underlying variables.

![picture]()

## Problems with Carnegie System
The process used by the Carnegie Classifications illustrates several problems:
+ Snapshot Data
+ 
+ 
+

@@

Structural Equation Models (SEM)

@@

## STEM and Non-STEM Factors
A more intuitive method would be to consider two latent factors:

@@

## Determining Group Membership
The SEM model returns a single factor score for each university. These can be used as inputs to a

## Group Membership

@@

## Conclusions and

>>>>

---
title: "An Alternative to the Carnegie Classifications:"
subtitle: "Using Structural Equation Models to Identify Similar Doctoral Institutions"
author: |
  | Paul Harmon
  | with Sarah McKnight, Laura Hildreth, Ian Godwin and Mark Greenwood
  | Montana State University

date: "April 2, 2018"
output:
  beamer_presentation:
    theme: "boxes"
    colortheme: "crane"
    fonttheme: "structuresmallcapsserif"
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```

```{r, include = FALSE}

#read in data
setwd("C:/Users/paulh/Documents/Carnegie-SEM/Carnegie_SEM")
cc2015 <- read.csv("data/CC2015data.csv",header = TRUE)

########2015################
cc2015.full <- read.csv("data/CC2015data.csv", header = TRUE, as.is = TRUE)
#updated file
#cc2015.full <- read.csv("Updated2015.csv", header = TRUE)

cc2015 <- cc2015.full[(cc2015.full$BASIC2015>14&cc2015.full$BASIC2015<18),]
cc2015$BASIC2015 <- factor(cc2015$BASIC2015)

#function for ranking the data
minrank <- function(x){rank(x, ties.method = "min")}

#dataset that we want to use
cc2015Ps<-
  na.omit(cc2015[,c("NAME","BASIC2010","BASIC2015","FACNUM","HUM_RSD",
                    "OTHER_RSD","SOCSC_RSD","STEM_RSD","PDNFRSTAFF","S.ER.D","NONS.ER.D")])

#calculate the ranked data
cc2015.r <- data.frame(cc2015Ps[,1:3],sapply(cc2015Ps[,-c(1:3)],minrank))

cc2015percap <- cc2015Ps[,c("PDNFRSTAFF","S.ER.D","NONS.ER.D")]/cc2015Ps$FACNUM

colnames(cc2015percap) <- c("PDNRSTAFF_PC", "S.ER.D_PC", "NONS.ER.D_PC")
#for Corrplot, nothing else

cc2015percap.r<-data.frame(sapply(cc2015percap,minrank))
cc2015_r <- cbind(cc2015.r, cc2015percap.r)

model4 <- '
HUMANITIES=~HUM_RSD+OTHER_RSD+SOCSC_RSD+NONS.ER.D+FACNUM
STEM=~STEM_RSD+PDNFRSTAFF+S.ER.D+FACNUM
Aggregate=~HUMANITIES+STEM'

library(lavaan)
lavaan_sem_new <- lavaan::sem(model4, data=cc2015_r, std.lv=TRUE,
                              orthogonal=FALSE, se="robust.huber.white")
lavaan::summary(lavaan_sem_new, standardized=TRUE, fit.measures=TRUE)
CCScores <- as.data.frame(lavaan::predict(lavaan_sem_new))
rownames(CCScores) <- cc2015Ps$NAME

#checks the standardized factor loadings
inspect(lavaan_sem_new, what = 'std')$lambda

library(mclust)
mcres<-Mclust(CCScores$Aggregate)
summary(mcres)

Classifications <- as.data.frame(mcres[14:15])
rownames(Classifications) <- cc2015Ps$NAME

table(mcres$classification, cc2015Ps$BASIC2015) ## pretty good!!

#Or more directly:
dens<-densityMclust(CCScores$Aggregate)
summary(dens)
par(mfrow=c(1,2))
plot(dens,what="diagnostic")

#Or clustering the STEM and HUMANITIES scores (bivariate version)
mcres2<-Mclust(CCScores[,1:2])

```

@@

The Carnegie Classifications delineate doctoral-granting institutions into three groups:

+ R1: Very High Research
+ R2: High Research
+ R3: Moderate Research

@@

## The Carnegie Classifications
![](figures/2015classifications.png)

## What Data are used?
Data used in the Carnegie Classifications are based on a snapshot from the Integrated Postsecondary Education Data System (IPEDS). The variables used in the classifications are:

+ **STEM expenditures** (in thousands of dollars)
+ **Non-STEM expenditures** (in thousands of dollars)
+ **Nontenurable Research Staff size**
+ STEM PhD Counts
+ Humanities PhD Counts
+ Social Science PhD Counts
+ Other PhD Counts

+ *Tenured/Tenure Track Faculty Headcount*

**Per-Capita**: The 3 variables in bold are used in per-capita variables by dividing by Faculty Headcount.

@@

The Carnegie Classifications are built on the following methodology:

+ **Rank** Institutions on 7 aggregate variables and 3 per-capita variables
+ **Index Creation**: PCA of aggregate and per-capita variables
+ **Plot the Indices**:  (per-capita vs aggregate)
+ **Create groups**:  (via line-drawing)

## Principal Component Analysis 
Principal Componenent Analysis is a method for dimension reduction that takes a set of p predictor variables and decomposes it into k principal components that explain the most variation in the underlying variables.

+ Principal Components are weighted averages of the observed variables
+ In the Carnegie Classifications, they create the aggregate index by taking the first principal component from a PCA of 7 aggregate variables
+ The per-capita index is the first principal component from a PCA of 3 per-capita variables

## Problems with Carnegie System
The process used by the Carnegie Classifications illustrates several problems:

+ **Based on Snapshot Data**: Data are based on a single snapshot rather than averages during the five-year period.  
+ **Changing Loadings**: Loadings can change yearly based on variability in the data. This has the potential to change which variables are most important in determining the aggregate and per-capita indices.  
+ **Dependence on Correlation**:  
+ **Group Determination**: The group determination is completely subjective. Lines are drawn after visual inspection of the plot.

This causes problems for schools that try to direct policy based on these classifications because the most important variables can be different from release to release based on variability in the data.

@@

Structural Equation Models (SEM) are used to model simultaneous equations, the use of latent or unobserved variables, and variables to be measured with error.

+ **Latent Variable Model**: Measures unobserved variables. 
+ **Measurement Model**: Relates the latent variables we are interested in to the manifest that we actually observe.

@@

We can think of the Carnegie Classifications in a latent modeling framework using this path diagram:

![Carnegie Model](analysis/carnegiepng.png)

##Problems with the Carnegie Method for SEM:
The variables used to measure the aggregate and per-capita latent variables are the same (with a slight per-capita transformation). The SEM cannot handle this level of correlation in the two latent factors; it **does not converge** and thus paramaters cannot be estimated.

## STEM and Non-STEM Factors: An Alternative
A more intuitive method would be to consider two latent factors:

@@

## STEM and Non-STEM Manifest Correlations

![](corrplot_numbers.png)

```{r, fig.cap = 'Correlations of Manifest Variables', fig.width =4 , fig.height = 2, eval = FALSE}
#RUN THIS FOR CORRPLOT: (commented out otherwise)
cc2015_r_new <- cc2015_r
names(cc2015_r_new) <- c("Name","2010 Basic","2015 Basic","Faculty Size", "Hum PhD",
                     "Other PhD","Soc. Sci PhD", "STEM PhD","Res. Staff", "STEM Exp", 
                     "NS Exp", "Res. Staff PC", "STEM Exp PC  ", "NS Exp PC")

cc2015_matrix2 <- as.matrix(cc2015_r_new[-c(1:3)])
corrmatrix <- Hmisc::rcorr(cc2015_matrix2)

par(mar = c(1,.8,.8,2))
corrplot::corrplot(corrmatrix$r, order="hclust", type = "upper", method = 'number')
text(-4,3, 'Correlation Plot', font = 2)
text(-4,2,'of Manifest Variables', font = 2)
rect(-7,1.3,-1,3.5)

```

##Proposed Model: STEM and Non-STEM
![Carnegie Model](analysis/onyx_final.png)

## Determining Group Membership
The SEM model returns a single factor-of-factor score for each university. These can be used as inputs to a clustering algorithm to determine both the optimal number of clusters and the optimal cluster membership.

__Potential Problems__:

+ Optimal Number of Clusters may be too large/small: We can fix this to a reasonable number if necessary (results here are fixed at 3). 
+ There are many different clustering methods to use (hierarchical clustering, mixture-model based methods, etc.).

## Group Membership with Mixture Model:

Using a mixture model, we can objectively define clusters. Additionally, we can illustrate uncertainty in classification for schools near the boundary.

```{r plot_classifications, fig.align = 'center',fig.height = 6, fig.width = 6}
Classifications <- mcres$classification
#creates a plot and colors by Carnegie Classification Colors

#adds the bobcat to plot
library(png)
library(grid)
img <- readPNG("msu_logo.png")
stf <- readPNG("stanford_trans.png")
g <- rasterGrob(img, interpolate=FALSE)
stan <- rasterGrob(stf, interpolate = FALSE)

a <- which(rownames(CCScores)=="Montana State University")
b <- which(rownames(CCScores) == 'Stanford University')

MSU_x <- CCScores$STEM[a]
MSU_y <- CCScores$HUMANITIES[a]
stan_x <- CCScores$STEM[b]
stan_y <- CCScores$HUMANITIES[b]
    ggplot(CCScores) + geom_point(aes(x = STEM, y = HUMANITIES, color = factor(Classifications)))+ 
      ggtitle("Predicted vs Actual Classifications") + theme_bw() + coord_fixed(ratio = 1)+
      theme_classic() + 
      #theme(legend.position="right", legend.box = "vertical") + 
      labs(color = "Classification") + 
       annotation_custom(g, xmin = MSU_x-.25 , xmax=MSU_x+.35, ymin= MSU_y -.3, ymax=MSU_y + .25) + annotation_custom(stan, xmin = stan_x-.25 , xmax=stan_x+.35, ymin= stan_y -.3, ymax=stan_y + .25)

```

## Uncertainty in Classifications
The Carnegie Classifications treat each classification as a fixed value, without accounting for the uncertainty inherent in the process.

+ Variability in underlying data
+ Uncertainty in PCA
+ Uncertainty in Mixture Modeling

Our method can be used to visualize differences in the data.

## Uncertainty Plots
```{r, fig.width = 8, fig.height = 5, fig.align = 'center'}

library(mclust)
mcres<-Mclust(CCScores$Aggregate)
#summary(mcres)
par(mfrow=c(1,2))
par(mar = c(1,1,1.2,1))
#plot(mcres,what="BIC")
plot(mcres,what="classification", col = c("red","green",'blue'))
legend('topleft',fill = c('red','green','blue'),legend = c("1",'2','3'))
plot(mcres,what="uncertainty", col = c("red",'green','blue') )
legend('topleft',fill = c('red','green','blue'),legend = c("1",'2','3'))
#plot(mcres,what="density")
```

## Shiny Applications: 
We developed shiny applications to assess sensitivity of both the Carnegie Classifications and the proposed SEM-Classifications for any institution of your choosing.

+ Sliders allow for user to change counts of PhDs, Staff/Faculty size, Expenditures
+ User can select any Doctoral-granting institution
+ The SEM application uses a fixed number of three clusters

These applications can be found at: paulharmon.shiny.io/SEM-class

## Shiny Applications:

![Screenshot of the Application](analysis/app_screenshot_small.png)

## Conclusions

The SEM-based model allows for several benefits compared to the Carnegie Classifications:

+ Latent variables do not have to be orthogonal like PCA-based indices
+ Latent variable modeling makes more sense than dimension-reduction (especially since we only have 7 variables)
+ Single-factor scores allow for comparison on a single dimension rather than using two scores.

@@

Future work will focus on determining an optimal clustering strategy. The Carnegie Classifications themselves are changing, so if they change the variables that are used, we will investigate development of either a new latent trait or sensitivity of additional manifest variables to measure the STEM and non-STEM latent traits.

## Questions:
\centering
Thank you!

### 2018-04-11 analysis/April_Presentation.Rmd
adding a ton of stuff to the repo

---
title: "An Alternative to the Carnegie Classifications:"
subtitle: "Using Structural Equation Models to Identify Similar Doctoral Institutions"
author: |
  | Paul Harmon
  | with Sarah McKnight, Laura Hildreth, Ian Godwin and Mark Greenwood
  | Montana State University

date: "April 2, 2018"
output:
  beamer_presentation:
    theme: "boxes"
    colortheme: "crane"
    fonttheme: "structuresmallcapsserif"
    fig_caption: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(ggplot2)
```

@@

#read in data
setwd("C:/Users/paulh/Documents/Carnegie-SEM/Carnegie_SEM")
cc2015 <- read.csv("data/CC2015data.csv",header = TRUE)

########2015################
cc2015.full <- read.csv("data/CC2015data.csv", header = TRUE, as.is = TRUE)
#updated file
#cc2015.full <- read.csv("Updated2015.csv", header = TRUE)

@@

cc2015percap.r<-data.frame(sapply(cc2015percap,minrank))
cc2015_r <- cbind(cc2015.r, cc2015percap.r)

@@

@@

@@

The Carnegie Classifications delineate doctoral-granting institutions into three groups:

+ R1: Very High Research
+ R2: High Research
+ R3: Moderate Research

@@

## The Carnegie Classifications
![](figures/2015classifications.png)

@@

+ **STEM expenditures** (in thousands of dollars)
+ **Non-STEM expenditures** (in thousands of dollars)
+ **Nontenurable Research Staff size**
+ STEM PhD Counts
+ Humanities PhD Counts
+ Social Science PhD Counts
+ Other PhD Counts

+ *Tenured/Tenure Track Faculty Headcount*

**Per-Capita**: The 3 variables in bold are used in per-capita variables by dividing by Faculty Headcount.

@@

The Carnegie Classifications are built on the following methodology:

+ **Rank** Institutions on 7 aggregate variables and 3 per-capita variables
+ **Index Creation**: PCA of aggregate and per-capita variables
+ **Plot the Indices**:  (per-capita vs aggregate)
+ **Create groups**:  (via line-drawing)

## Principal Component Analysis 
Principal Componenent Analysis is a method for dimension reduction that takes a set of p predictor variables and decomposes it into k principal components that explain the most variation in the underlying variables.

+ Principal Components are weighted averages of the observed variables
+ In the Carnegie Classifications, they create the aggregate index by taking the first principal component from a PCA of 7 aggregate variables
+ The per-capita index is the first principal component from a PCA of 3 per-capita variables

## Problems with Carnegie System
The process used by the Carnegie Classifications illustrates several problems:

+ **Based on Snapshot Data**: Data are based on a single snapshot rather than averages during the five-year period.  
+ **Changing Loadings**: Loadings can change yearly based on variability in the data. This has the potential to change which variables are most important in determining the aggregate and per-capita indices.  
+ **Dependence on Correlation**:  
+ **Group Determination**: The group determination is completely subjective. Lines are drawn after visual inspection of the plot.

This causes problems for schools that try to direct policy based on these classifications because the most important variables can be different from release to release based on variability in the data.

@@

![Carnegie Model](analysis/carnegiepng.png)

##Problems with the Carnegie Method for SEM:
The variables used to measure the aggregate and per-capita latent variables are the same (with a slight per-capita transformation). The SEM cannot handle this level of correlation in the two latent factors; it **does not converge** and thus paramaters cannot be estimated.

@@

## STEM and Non-STEM Manifest Correlations

![](corrplot_numbers.png)

```{r, fig.cap = 'Correlations of Manifest Variables', fig.width =4 , fig.height = 2, eval = FALSE}
#RUN THIS FOR CORRPLOT: (commented out otherwise)
cc2015_r_new <- cc2015_r
names(cc2015_r_new) <- c("Name","2010 Basic","2015 Basic","Faculty Size", "Hum PhD",
                     "Other PhD","Soc. Sci PhD", "STEM PhD","Res. Staff", "STEM Exp", 
                     "NS Exp", "Res. Staff PC", "STEM Exp PC  ", "NS Exp PC")

cc2015_matrix2 <- as.matrix(cc2015_r_new[-c(1:3)])
corrmatrix <- Hmisc::rcorr(cc2015_matrix2)

##Proposed Model: STEM and Non-STEM
![Carnegie Model](analysis/onyx_final.png)

@@

+ Optimal Number of Clusters may be too large/small: We can fix this to a reasonable number if necessary (results here are fixed at 3). 
+ There are many different clustering methods to use (hierarchical clustering, mixture-model based methods, etc.).

@@

Using a mixture model, we can objectively define clusters. Additionally, we can illustrate uncertainty in classification for schools near the boundary.

```{r plot_classifications, fig.align = 'center',fig.height = 6, fig.width = 6}
Classifications <- mcres$classification
#creates a plot and colors by Carnegie Classification Colors

#adds the bobcat to plot
library(png)
library(grid)
img <- readPNG("msu_logo.png")
stf <- readPNG("stanford_trans.png")
g <- rasterGrob(img, interpolate=FALSE)
stan <- rasterGrob(stf, interpolate = FALSE)

a <- which(rownames(CCScores)=="Montana State University")
b <- which(rownames(CCScores) == 'Stanford University')

MSU_x <- CCScores$STEM[a]
MSU_y <- CCScores$HUMANITIES[a]
stan_x <- CCScores$STEM[b]
stan_y <- CCScores$HUMANITIES[b]
    ggplot(CCScores) + geom_point(aes(x = STEM, y = HUMANITIES, color = factor(Classifications)))+ 
      ggtitle("Predicted vs Actual Classifications") + theme_bw() + coord_fixed(ratio = 1)+
      theme_classic() + 
      #theme(legend.position="right", legend.box = "vertical") + 
      labs(color = "Classification") + 
       annotation_custom(g, xmin = MSU_x-.25 , xmax=MSU_x+.35, ymin= MSU_y -.3, ymax=MSU_y + .25) + annotation_custom(stan, xmin = stan_x-.25 , xmax=stan_x+.35, ymin= stan_y -.3, ymax=stan_y + .25)

@@

## Uncertainty in Classifications
The Carnegie Classifications treat each classification as a fixed value, without accounting for the uncertainty inherent in the process.

+ Variability in underlying data
+ Uncertainty in PCA
+ Uncertainty in Mixture Modeling

Our method can be used to visualize differences in the data.

## Uncertainty Plots
```{r, fig.width = 8, fig.height = 5, fig.align = 'center'}

library(mclust)
mcres<-Mclust(CCScores$Aggregate)
#summary(mcres)
par(mfrow=c(1,2))
par(mar = c(1,1,1.2,1))
#plot(mcres,what="BIC")
plot(mcres,what="classification", col = c("red","green",'blue'))
legend('topleft',fill = c('red','green','blue'),legend = c("1",'2','3'))
plot(mcres,what="uncertainty", col = c("red",'green','blue') )
legend('topleft',fill = c('red','green','blue'),legend = c("1",'2','3'))
#plot(mcres,what="density")
```

@@

+ Sliders allow for user to change counts of PhDs, Staff/Faculty size, Expenditures
+ User can select any Doctoral-granting institution
+ The SEM application uses a fixed number of three clusters

These applications can be found at: paulharmon.shiny.io/SEM-class

## Shiny Applications:

![Screenshot of the Application](analysis/app_screenshot_small.png)

@@

## Questions:
\centering
Thank you!

>>>>

---
title: "Untitled"
author: "Paul Harmon"
date: "April 2, 2018"
output: beamer_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

@@

#read in data
#setwd("~/Carnegie-SEM/data")
cc2015 <- read.csv("CC2015data.csv",header = TRUE)

########2015################
cc2015.full <- read.csv("CC2015data.csv", header = TRUE, as.is = TRUE)
#updated file
#cc2015.full <- read.csv("Updated2015.csv", header = TRUE)

@@

cc2015percap.r<-data.frame(sapply(cc2015percap,minrank))
cc2015_r <- cbind(cc2015.r, cc2015percap.r)
#RUN THIS FOR CORRPLOT: (commented out otherwise)
names(cc2015_r) <- c("Name","2010 Basic","2015 Basic","Faculty Size", "Hum PhD",
                     "Other PhD","Soc. Sci PhD", "STEM PhD","Res. Staff", "STEM Exp", 
                     "NS Exp", "Res. Staff PC", "STEM Exp PC  ", "NS Exp PC")

cc2015_matrix2 <- as.matrix(cc2015_r[-c(1:3)])
corrmatrix <- Hmisc::rcorr(cc2015_matrix2)

@@

model4 <- '
non-STEM = ~ Hum PhD + Other PhD + Soc. Sci PhD + 
STEM
'

@@

#path diagram

source('https://raw.githubusercontent.com/brandmaier/onyxR/master/tools/install.R')
library(onyxR)
onyx(lavaan_sem_new)

library(beanplot)
beanplot(CCScores$Aggregate,log="",col="yellow",method="jitter")

@@

The Carnegie Classifications delineate doctoral-granting institutions into three groups: 
+ R1: Very High Research
+ R2: High Research
+ R3: Moderate Research

@@

@@

+ STEM expenditures (in thousands of dollars)
+ Non-STEM expenditures (in thousands of dollars)
+ STEM PhD Counts
+ Humanities PhD Counts
+ Social Science PhD Counts
+ Nontenurable Research Staff size
+ Tenured/Tenure Track Faculty Headcount

@@

+ **Rank** Institutions
+ **Index Creation**: PCA of aggregate and per-capita variables
+ **Plot the Indices**:  (per-capita vs aggregate)
+ **Create groups**:  (via line-drawing)

## Plot of the Carnegie Classifications
```{r}

## Principal Components Analysis 
Principal Componenents Analysis is a method for dimension reduction that takes a set of p predictor variables and decomposes it into k principal components that explain the most variation in the underlying variables.

![picture]()

## Problems with Carnegie System
The process used by the Carnegie Classifications illustrates several problems:
+ **Based on Snapshot Data**: Data are based on a single snapshot rather than averages during the five-year period.  
+ **Changing Loadings**: Loadings can change yearly based on variability in the data. This has the potential to change which variables are most important in determining the aggregate and per-capita indices.  
+ **Dependence on Correlation**:  
+ **Group Determination**: The group determination is completely subjective. Lines are drawn after visual inspection of the plot ().

\\
This causes problems for schools that try to direct policy based on these classifications because the most important variables can be different from release to release based on variability in the data.

@@

![Carnegie Model](carnegiepng.png)

##Problems with the Carnegie Method for SEM:
The variables used to measure the aggregate and per-capita latent variables are literally the same (with a slight per-capita transformation). The SEM cannot handle this level of correlation in the two latent factors; it **does not converge** and thus paramaters cannot be estimated.

@@

![Correlation Plot of Observed Variables](corrplot_numbers.png)

##Proposed Model: STEM and Non-STEM
![Carnegie Model](onyx_final.png)

@@

+ Optimal Number of Clusters may be too large/small: We can fix this to a reasonable number if necessary (results here are fixed at 3). 
+ There are many different clustering methods to use (hierarchical clustering, mixture-model based methods, etc. )

@@

Using a mixture model approach, we can obtain a solution with reasonable clusters. Additionally, we can account for/illustrate uncertainty in classification for schools near the boundary.

```{r plot_classifications}
Classifications <- mcres$classification
#creates a plot and colors by Carnegie Classification Colors  
    ggplot(CCScores_r_cov_new) + geom_point(aes(x = STEM, y = HUM, color = factor(Classifications), shape = factor(symbols), size = factor(symbols)))+ 
      ggtitle("Predicted vs Actual Classifications") + theme_bw() + coord_fixed(ratio = 1)+
      theme_classic() + 
      #theme(legend.position="right", legend.box = "vertical") + 
      labs(color = "Classification")

@@

@@

![Screenshot of the Application](app_screenshot.png)

@@

##Questions:
![]()
